---
title: "Tutorial: Specific plots in ggplot"
subtitle: "Research Practicum in CSS"
date: today
toc: true
author: "Renata Topinkova"
engine: knitr
format: live-html
webr:
  packages:
    - dplyr
    - ggplot2
    - gapminder
    - forcats
  render-df: gt-interactive
---

{{< include ./_extensions/r-wasm/live/_knitr.qmd >}}

::: callout-warning
# This is an **experimental** version of the tutorial.

If you spot any errors, please let me know in the forum, so I can fix it.
:::

## Loading packages and data

First, load the `tidyverse` and `gapminder` packages

```{r}
#| message: false
library(dplyr)
library(ggplot2)
library(forcats)
library(gapminder)
```

::: {.callout-note collapse="true"}
# What are these packages for?

-   **`dplyr`**: data manipulation and transformation (e.g., filtering, summarising, changing columns).

-   **`ggplot2`**: data visualizations.

-   **`forcats`**: handling and reordering of categorical variables (factors).

-   **`gapminder`**: provides a dataset of global economic and demographic trends for practice and exploration. More on gapminder project [here](https://www.gapminder.org/).
:::

## Explore the data

Examine the `gapminder` object.

```{webr}
gapminder
```

Take a `glimpse()` into its contents:

```{webr}
#| exercise: ex_glimpse

_____(____)
```

:::{.hint exercise="ex_glimpse"}
:::{.callout-note collapse="false"}

## Hint 
```r 
glimpse(nameofdataset)
``` 

:::
:::

:::{.solution exercise="ex_glimpse"}
:::{.callout-note collapse="false"}

## Solution 

```{r}
glimpse(gapminder)
```


:::
:::


## 1. Bar charts

Now, let's take a look on how many observations per continent we have in the dataset.

Remember that every ggplot call starts with:

``` r
ggplot(data = data, mapping = aes(___)) +
   geom_*() +
   additional calls
```

Since we want to make a bar chart, there are two `geom_*`s we can choose from. Either `geom_bar()` or `geom_col()`. They both serve the same function but require different `aes`thetics.

::: callout-note
# `geom_bar()` vs `geom_col()` ?

-   `geom_bar()` requires specifying only what goes into one of the axes (`x` OR `y`) of the plot. It calculates the frequency of each category for you for the number of rows per category in your data.

-   `geom_col()` requires specifying both `x` and `y` axis. This means that your dataset should already have a column with frequencies (or other metric that you want to show, such as percentage, rate, or median).
:::

Let's say that we want to quickly check the number of observations per continent. Because our data does not have any column that would indicate the frequency of each continent, we need to use `geom_bar()` to calculate it for us.

```{webr}
ggplot(gapminder, aes(x = continent)) + 
  geom_bar()
```

:::{.callout-note collapse=true}
## Solution with `geom_col()`

As we mentioned before, `geom_col()` expects we already have a column with frequencies or other values we want to plot. We can get our data into the required shape by calling `count()`
```{r}
gapminder |> 
  count(continent)
```


Now that the data has a new column called `n`, which holds the counts of rows per continent, we can use `geom_col()`:

```{webr}
gapminder |> 
  count(continent) |> 
  ggplot(aes(x = continent, y = n)) + 
  geom_col()
```
:::

### Horizontal bar chart

To make our plot horizontal, we can make a quick adjustment. Replace the \_\_\_ with `y`. This will **map** the continent column to the `y` axis instead of the x axis (which we have done above) and the counts to the x axis.

#### 🧠 Try it out!

```{webr}
#| exercise: ex_horizontal
ggplot(gapminder, aes(__ = continent)) + 
  geom_bar()
```


::: { .hint exercise="ex_horizontal"}
::: { .callout-note collapse="false"}
## Hint 

Replace the `_` with `y`.

:::
:::


:::{.solution exercise="ex_horizontal"}

::: { .callout-note collapse="false"}

```{r}
ggplot(gapminder, aes(y = continent)) + 
  geom_bar()
```

:::
:::


### Tricky parts in "easy" plots - reordering


Notice that our plot is ordered alphabeticaly - this is the default ordering in ggplot. In vertical bar charts, bars are ordered alphabetically from left to right. In horizontal bar charts, bars are ordered alphabetically from top to bottom. 

```{r}
ggplot(gapminder, aes(x = continent)) + 
  geom_bar()
```

What if we wanted to change the order of bars based on their counts? We can do this with the use of the `forcats` package.

If we call `glimpse()` on our dataset, we can see that `continent` has a class of `fct`. This means that it is a factor variable. A factor variable in R is used to represent **categorical data with a fixed number of unique values, known as levels**.

```{webr}
glimpse(gapminder)
```


We can see the levels of the `continent` column by calling the `levels` function:

```{webr}
levels(gapminder$continent)
```


#### Order by frequency

:::{.callout-warning}

## Should I always reorder bars based on frequency?

**No!** You should only reorder your bars if they represent **nominal variables**, which have no inherent order (e.g., types of fruits, names of countries). Reordering nominal variables by frequency or another metric can help make your data easier to interpret.

However, if your bars represent **ordinal variables** — variables with a meaningful, inherent order (e.g., Likert scales such as Very Good > Good > Bad > Very Bad)—the original order should always be preserved, even if it doesn’t follow the category’s frequency. 
:::


To order our bars by frequency, we can call the `fct_infreq()` function on the `continent` column. 

```{webr}
gapminder |> 
  mutate(continent = fct_infreq(continent)) |> 
  ggplot(aes(x = continent)) + 
  geom_bar()
```


:::{.callout-note collapse=true}

## Do I have to call `fct_infreq()` in `mutate()` before plotting? 

Not necessarily. The code will still work if you do the reordering directly in `ggplot`. Try it out! 

```{webr}
gapminder |> 
  ggplot(aes(x = fct_infreq(continent))) + 
  geom_bar()
```

:::

#### Reverse order

To reverse the order of the previous bar from descending to ascending, we can call `fct_rev()`. Note that you need to first order your bars based on frequency with `fct_infreq()` and only after that you can reverse the order! 

```{webr}
gapminder |> 
  mutate(continent = fct_rev(fct_infreq(continent))) |> 
  ggplot(aes(x = continent)) + 
  geom_bar()
```
:::{.callout-note collapse=true}
## Why does the order of `fct_*` functions matter? 

Let's see what happens if the order is reversed and we first reverse the order, and then sort it by frequency. 

We start with this plot: 

```{r}
gapminder |> 
    ggplot(aes(x = continent)) + 
    geom_bar()
```

In the second step, we reverse the order of continent:

```{r}
gapminder |> 
    mutate(continent = fct_rev(continent)) |> 
    ggplot(aes(x = continent)) + 
    geom_bar()
```

In the last step, we sort the bars by frequency:

```{r}
gapminder |> 
    mutate(continent = fct_infreq(fct_rev(continent))) |> 
    ggplot(aes(x = continent)) + 
    geom_bar()
```

This way, we would end up with the bar chart in descending order again! 

:::




#### Move things around

Lastly, we may want to move some bars around. In this case, it does not make a lot of sense but it may come handy - we can also move our bars manually. In this case, we can move "Europe" to the front. 


```{webr}
gapminder |> 
    mutate(continent = fct_relevel(continent, "Europe")) |> 
    ggplot(aes(x = continent)) + 
    geom_bar()
```


We can also move more than one bar at a time. This can be done by adding another category to the `fct_relevel()` function. 

#### 🧠 Try it out!
Move "Oceania" next to "Europe".

```{webr}
#| exercise: ex_oc
gapminder |> 
    mutate(continent = fct_relevel(continent, "Europe", ____)) |> 
    ggplot(aes(x = continent)) + 
    geom_bar()
```



:::{.hint exercise="ex_oc"}
:::{.callout-note collapse="false"}

## Hint 


```r
gapminder |> 
    mutate(continent = fct_relevel(continent, "Europe", "nameofcategory")) |> 
    ggplot(aes(x = continent)) + 
    geom_bar()
```
:::
:::


:::{.solution exercise="ex_oc"}
:::{.callout-note collapse="false"}

## Solution

```{r}
gapminder |> 
    mutate(continent = fct_relevel(continent, "Europe", "Oceania")) |> 
    ggplot(aes(x = continent)) + 
    geom_bar()
```


:::
:::



### Changing colors

Change the color of all the bars to `cyan4`. In the previous tutorial, we worked with a scatterplot. When we wanted to change the color of the dots in our plot, we used `color` to do this. 

```{webr}
ggplot(gapminder, aes(x = continent)) + 
  geom_bar(color = "cyan4")
```

This is not what we expected! What happened? In ggplot2, the `color` aesthetic applies to the **outlines of shapes**, while `fill` applies to the **insides of shapes**. In our case, the `color = "cyan4"` created a thin outline around the bars.


:::{.callout-note}
## Color vs fill?

-   For points (e.g., in scatterplots), the shapes typically have only an outline, so `color` is used to define their appearance.

-  For bars (e.g., in bar plots), the interior is the primary visible feature, so `fill` is used to color them.

:::


#### 🧠 Try it yourself

Knowing the difference in how R treats outlines and interiors of geoms, change the code so the bars are filled in the `cyan4` color.

```{webr}
#| exercise: ex_fill
ggplot(gapminder, aes(x = continent)) + 
  geom_bar(___ = ____)
```

:::{.hint exercise="ex_fill"}
:::{.callout-note collapse="false"}

## Hint
```r 
geom_bar(fill = ____)

```
:::
:::

:::{.solution exercise="ex_fill"}
:::{.callout-note collapse="false"}

## Solution

```{r}
ggplot(gapminder, aes(x = continent)) + 
  geom_bar(fill = "cyan4")
```


:::
:::


:::{.callout-tip collapse=true}

## Quick note on colors in R

There are more than 650 pre-defined named colors in R, such as various shades of red, blue, cyan, coral, etc.! You can see some of them [here](https://sape.inf.usi.ch/sites/default/files/ggplot2-colour-names.png). 

Additionally, R comes with multiple color pallettes: 

```{r}
#| fig-height: 7
RColorBrewer::display.brewer.all()
```


Alternatively, you can also use Hex codes to get exactly the color you want - see more [here](https://htmlcolorcodes.com/).

:::

### Label the plot

#### 🧠 Try it yourself

Add labels (title, x and y axis) to the plot.

```{webr}
#| exercise: ex_lab
ggplot(gapminder, aes(x = continent)) + 
  geom_bar(fill = "cyan4") +
  ____
```

:::{.hint exercise="ex_lab"}
:::{.callout-note collapse="false"}

## Hint

Use the `labs()` function. 

If you're unsure how to proceed, consult the first tutorial. 

:::
:::

:::{.solution exercise="ex_lab"}
:::{.callout-note collapse="false"}

## Solution

```{r}
ggplot(gapminder, aes(x = continent)) + 
  geom_bar(fill = "cyan4") +
  labs(title = "Number of observations per continent", 
       x = "Continent",
       y = "Count") 
```


:::
:::



## 2. Histograms and density plots

A **histogram** is a plot that shows the distribution of a **continuous variable** by dividing it into **intervals (bins)** and counting how many observations fall into each bin. Unlike bar plots, which display counts or values for categorical variables, histograms represent frequencies or densities of continuous data.


### 🧠 Try it yourself: Data wrangling

Filter the data, so we only work with data from year 1952

```{webr}
#| exercise: ex_filter
gapminder_filtered <- gapminder |> 
  _____________
```


:::{.hint exercise="ex_filter"}
:::{.callout-note collapse="false"}

## Hint

Use the `filter()` function from the `dplyr` package.

:::
:::

:::{.solution exercise="ex_filter"}
:::{.callout-note collapse="false"}

## Solution

```{r}
gapminder_filtered <- gapminder |> 
  filter(year == 1952)
```


:::
:::


Make a histogram of `lifeExp` using the data you have created in the previous step. For histogram, we need only one `aes`thetic - the variable that will be plotted, in our case, `lifeExp`. 


```{webr}
#| setup: true
#| exercise:
#|  - ex_hist
#|  - ex_hist_bin
gapminder_filtered <- gapminder |> 
  filter(year == 1952)
```

```{webr}
#| exercise: ex_hist
gapminder_filtered |> 
  ggplot(aes(x = lifeExp)) + 
  geom_histogram()
```

### Binwidths


An important feature in a histogram is its binwidth. Binwidth determines the width of each bin (interval) used to group data points. It plays a crucial role in how the data distribution is visualized:

-   Too small binwidth: Creates many narrow bins, leading to a noisy and overly detailed plot that can obscure overall patterns.

-   Too large binwidth: Uses fewer, broader bins, oversimplifying the data and potentially hiding important features of the distribution.

#### 🧠 Try it yourself

Play around with the `Select the binwidth` slider (below the histogram) to see how different binwidths change the interpretation.

```{webr}
#| exercise: ex_hist_bin
#| autorun: true
#| input:
#|   - n
gapminder_filtered |> 
  ggplot(aes(x = lifeExp)) + 
  geom_histogram(binwidth = n)
```

```{ojs}
//| echo: false
viewof n = Inputs.range([1, 30], {step: 1, label: "Select the binwidth"})
```



### Density plot

A density plot is a smoothed visualization of the distribution of a continuous variable, estimating its probability density function. Unlike histograms, which use bins, density plots provide a continuous curve, making it easier to see patterns and compare distributions.


#### 🧠 Try it yourself 

Change the code below from a histogram to a **density plot** (R: `geom_density()`.

```{webr}
gapminder_filtered |> 
  ggplot(aes(x = lifeExp)) + 
  geom_histogram() # change me to density plot!
```


## 3. Boxplots and violins

A boxplot is a graphical representation of the **distribution** of a dataset that highlights its key summary statistics: the median, quartiles, and potential outliers. It shows the spread and symmetry of the data, with the box indicating the interquartile range (IQR) and "whiskers" extending to data points within 1.5 times the IQR. Outliers are plotted as individual points.

In R, we can use the `geom_boxplot()` function.

```{webr}
gapminder |> 
  ggplot(aes(x = lifeExp)) + 
  geom_boxplot()

```

Seeing the distribution of life expectancy for all countries and years is nice, but it is not very informative. So let's improve our graph.

We can make boxplots of life expectancy by continents by mapping `continent` to one of the axes.

```{webr}
gapminder |> 
  ggplot(aes(x = lifeExp, y = continent)) + 
  geom_boxplot() 

```

### Reordering.. again! 

Sometimes, we may want to reorder categories in a plot, for example, arranging continents so that the one with the lowest median life expectancy appears first, and the one with the highest median appears last.

To achieve this, we can reorder the continent column with `fct_reorder()` before passing the data into ggplot. 

This involves specifying:

-   Which column to reorder: `continent`

-   The column to use for ordering: `lifeExp`

-   The statistic to base the ordering on: `.fun = median` (default for fct_reorder, so we could omit this part in the code below).

```{webr}
gapminder |>     
  mutate(continent = fct_reorder(continent, lifeExp, .fun = median)) |>    
  ggplot(aes(x = lifeExp, y = continent)) + 
  geom_boxplot()
```


### Outliers visibility
We can also change the appearance of outliers. Type `?geom_boxplot` into your console and find out how to change the size and opacity of outliers. Experiment with different options.

```{webr}
gapminder |>     
  mutate(continent = fct_reorder(continent, lifeExp, .fun = median)) |>    
  ggplot(aes(x = lifeExp, y = continent)) + 
  geom_boxplot(outlier.alpha = 0.25)
```

### Alternative to boxplots: Violin plots

A violin plot is a visualization that combines aspects of a boxplot and a density plot, showing both the summary statistics and the distribution of a continuous variable. 

```{webr}
gapminder |>     
  mutate(continent = fct_reorder(continent, lifeExp, .fun = median)) |>    
  ggplot(aes(x = lifeExp, y = continent)) + 
  geom_violin(draw_quantiles = c(0.25, 0.5, 0.75))
```
### BONUS: Adding observations to boxplots or violins

Sometimes, you may want to add individual observations on top of boxplots or violin plots to provide additional detail by showing the raw data points. This is often done using `geom_jitter()` (essentially a `geom_point` with added "noise" to prevent overplotting) or `geom_point()` (for direct plotting). Overlaying observations helps identify patterns, clusters, or outliers that may be hidden by summary statistics alone, offering a richer view of the data.

#### 🧠 Try it yourself

Experiment with the code below: You can e.g., change the `alpha` parameter or explore the difference between plotting the individual observations with `geom_point()` vs with `geom_jitter()`.

```{webr}
gapminder |>     
  mutate(continent = fct_reorder(continent, lifeExp, .fun = median)) |>    
  ggplot(aes(x = lifeExp, y = continent)) + 
  geom_boxplot(outlier.alpha = 0.25) + 
  geom_jitter(alpha = 0.2)
```


## 4. Line plots

A line plot is used to visualize the relationship between two variables, typically with one variable on the x-axis (often time) and another on the y-axis. Points are connected by lines, making it ideal for showing trends, changes, or patterns over time or across a continuous variable.

Let's make a new object that contains only data for Japan, Korea, Rep., Korea, Dem. Rep., and China, and call it `asia`.

```{webr}
#| autorun: true
asia <- gapminder |> 
  filter(country %in% c("Japan", "Korea, Rep.", "Korea, Dem. Rep.", "China"))

```

Following what we've learned before, we put year on x-axis, and gdp per capita on the y axis, and call `geom_line()` to make a line plot. 

```{webr}
asia |> 
  ggplot(aes(x = year, y = gdpPercap)) + 
  geom_line()
```

This does not seem right... What went wrong? We can investigate the relationship between year and gdp per capita with a scatterplot. What do you see?

```{webr}
asia |> 
  ggplot(aes(x = year, y = gdpPercap)) + 
  geom_point() 
```


The "seesawing" effect in the line plot occurs because `geom_line()` is connecting all the data points for multiple countries into a single line. By default, ggplot assumes all points belong to **one continuous series** unless explicitly told otherwise. In this case, it connects the GDP per capita values across all years for all four countries together, rather than treating each country as a separate series.

To fix this, we need to **group** the data by country so that each country has its own distinct line. This can be done by adding a `group` or `color` aesthetic.

```{webr}
asia |> 
  ggplot(aes(x = year, y = gdpPercap, color = country)) + 
  geom_line()
```


### Changing colors

`scale_color_manual()`is used to manually specify the colors for the levels of a categorical variable when using the `color` aesthetic (which is the case in our line plot as we mapped country to color).


```{webr}
asia |> 
  ggplot(aes(x = year, y = gdpPercap, color = country)) + 
  geom_line() + 
  scale_color_manual(values = c("Japan" = "gray", "Korea, Rep." = "blue", "Korea, Dem. Rep." = "black", "China" = "red"))
```

:::{.callout-tip collapse=true}
## scale_*_* functions

In the code above, we can use `scale_color_manual` because our variable country (which we want to color in a specific way), is mapped to the `color` aesthetic. 

However, if we wanted to change colors in a barplot, we would have to use `scale_fill_manual` as in this case, the bar colors (their interiors) are not mapped to `color` aesthetic but a `fill` aesthetic. 

Similarly, if our variable is mapped to `shape`, we would have to use `scale_shape_manual()`.


Additionally, we may want to change the units or labels on some of the axes. This would be done with `scale_x_continuous` for an x-axis with continous values, `scale_x_discrete` for an a-xis with discrete values, etc.

:::

## Adding text to plots

Adding text annotations to your data visualizations can enhance clarity and storytelling by highlighting key points, trends, or outliers. It helps viewers quickly grasp important insights without needing to analyze the plot in detail, making your visualizations more accessible and impactful.


One easy way to annotate our data can be adding direct labels. We can add labels to our data with `geom_text()`. 

`geom_text()` will by default inherit any aesthetic that is inside the`ggplot()` call. This specifies where the labels will be placed. We also need to specify where will the label (text) for our annotation come from. In our case, it comes from the `n` column.

```{webr}
gapminder |> 
  count(continent) |>
  ggplot(aes(x = continent, y = n)) +
  geom_col() +
  geom_text(aes(label = n))
  
```
This is a good start but we may want to do some additional changes. The text is awkwardly in the center of the bar. Maybe we want to raise it up or push it down. We can change this with `geom_text` argument `vjust` (vertical justification). The size of the text also seems a bit too big. We can change this by modifying the `size` argument. Both of those options are **set** for all observations. This means that we have to do it **outside** the `aes()` function.

🧠 Feel free to modify the code below to see the difference. You may also want to change the `color` of the text.

```{webr}
gapminder |> 
  count(continent) |>
  ggplot(aes(x = continent, y = n)) +
  geom_col() +
  geom_text(aes(label = n), 
            vjust = -0.2, 
            size = 3)
```

Don't forget to label your data!
```{webr}
gapminder |> 
  count(continent) |>
  ggplot(aes(x = continent, y = n)) +
  geom_col() +
  geom_text(aes(label = n), 
            vjust = -0.2, 
            size = 3) + 
  labs(title = "Number of observations per continent", 
         x = "Continent",
         y = "Count")
```



Another common approach, especially when working with line plots, is to highlight certain observations or time periods. 

We can start by drawing a vertical line to indicate where a change happened (`xintercept` marking which point on the axis will the line go through) with `geom_vline()`. We can also specify the type of line with `linetype`.

```{webr}
asia |> 
  ggplot(aes(x = year, y = gdpPercap, color = country)) + 
  geom_line() + 
  scale_color_manual(values = c("Japan" = "gray", "Korea, Rep." = "blue", "Korea, Dem. Rep." = "black", "China" = "red")) + 
  geom_vline(xintercept = 1970, linetype = "dotted")
```

Then, we can add more information with a text. If there is not many things to highlight, we can add it manually with `annotate()`. Note that the `x` argument in `annotate()` determines the **center** of the text.

🧠  Try to modify the code below to move the annotation around by changing the `x` and  `y`arguments. You can also try changing the `"text"` argument for `label` to see the difference.

```{webr}
asia |> 
  ggplot(aes(x = year, y = gdpPercap, color = country)) + 
  geom_line() + 
  scale_color_manual(values = c("Japan" = "gray", "Korea, Rep." = "blue", "Korea, Dem. Rep." = "black", "China" = "red")) + 
  geom_vline(xintercept = 1970, linetype = "dotted") + 
  annotate("text", x = 1978, y = 25000, label = "Something happened?")
```

